Investigation of Limited Detection Schemes for Light Scattering of Optically Trapped Asymmetric Particles

### 2022-12-21 Manuscript_rev2.tex
working on manuscript

\begin{highlights}
\item Asymmetric dimers undergo a full inversion in the presence of an potential well. This only occurs if the size ratio is greater than 1:2.  
\item Orientation can be classified by applying Bayesian inference to light intensity signals. 
\item The choice of detection angles has no noticeable effect on the performance of characterising orientation.
\item Performance of model is determined in part due to uncertainty in the measurement precision. 
\item Future applications in continuos processes where particle size is changing with time are being considered.  
\end{highlights}

@@

Since their invention in the late 1980's, Optical Tweezers have found application in experiments ranging from single molecule biophysics \cite{Bustamante2021Biophysics} to those that test the fundamental assumptions of Quantum Mechanics  \cite{yin2013large} thanks mainly to the ability of the tweezer to transduce and detect forces on the order of a few pico newtons. \textit{Further characterisation of the trapped entity would provide useful information, as explored in for example} metrological measurements \cite{arita2020coherent} and colloidal aggregation \cite{burns1990optical}. Chemical characterisation of the trapped entities has been achieved using spectroscopic techniques such as Raman Scattering \cite{gupta2014raman}, while dynamical characterisation has been demonstrated using a Quadrant Photo Detector,  by following the centre-of-mass Brownian motion of the trapped entity  \cite{friedrich2012tuning} and measuring rotation  of the centre-of-mass \cite{yifat2021facile}.

@@

\subsection{Brownian Simulation}
\label{sec:brownian}

We use the Brownian OT package developed by Fung~\textit{et~al}. \cite{Vigilante2020Brownian_OT} to simulate the motion of a asymmetric dimer within an optical trap. Brownian OT combines MSTM \cite{Mishchenko1996MSTM} and ``Optical Tweezer Toolbox'' (\textit{ott}) \cite{Lenton2020} to simulate the motion of arbitrary
shaped sphere clusters. We simulated the motion of a silica dimer
trapped in a highly focused Gaussian beam.

The dimer's Brownian motion is simulated by calculating its diffusion tensor according to Farsund's \emph{et al} work \cite{Farsund1996}. For computing the optical force Brownian OT uses the T-Matrix method to calculate scattering of the dimer. MSTM computes the T-matrix of the dimer, and then passes the T-matrix to \textit{ott} to compute the optical forces. The benefit of this approach is that MSTM can accurately compute the T-matrix for any arbitrary collection of spheres once, and then \textit{ott} can calculate the optical forces and torques. The total displacement is the sum of the optical and Brownian contributions.

Brownian OT returns the dimer's motion as a \textit{csv} file containing for each time step the $[x,y,z]$ position of the dimer's centre of mass relative to the trap focus, and the orientation represented as a quaternion.

@@

Our goal is to determine the orientation of our trapper dimer based on the scattering measurements $I(\hat{n}, \ \theta_k)$ collected from three detection fibres situated at $\theta_1,\ \theta_2, \ \theta3, \ \& \ \theta_4$. Rather than get an exact estimate of the dimer's orientation the possible orientation space is discretized into $\textit{n}_{ref} \ = \ 30$ possible reference orientations $\hat{\bf n}_{\alpha}$ that are evenly distributed on a unit sphere \cite{Rey2006}, see Figure~\ref{fig:dimer} b. The maximum spacing between two neighbouring reference orientations is roughly 0.895 radians.

Using MSTM we compute the raw intensities that would be collected from the dimer in each reference orientation $I(\hat{\bf n}_{\alpha}, \theta_k)$. We noticed that keeping all four angles in the same plane resulted in identical signals, to counter this we projected $\theta_4$ out of the horizontal plane and into the vertical plane ($\phi = 90$) resulting in unique signals for each reference orientation. 
%
These raw intensities are normalized according to
\begin{align}
\label{eq:scale}
  y_k(\hat{\bf n}_\alpha)
  = 
  \frac{I(\hat{\bf n}_\alpha, \theta_k) - \langle I(\hat{\bf n},\theta_k) \rangle } 
  {\langle I^2(\hat{\bf n},\theta_k) \rangle -\langle I(\hat{\bf n}, \theta_k)\rangle^2}
\end{align}
where the denominator is simply the standard deviation for our values of $I(\hat{\bf n}, \theta_k)$. The reference orientations, raw intensities, and scaled signals are given in Tables~\ref{tab:A1} and \ref{tab:A2}. The distribution of collected signals bares no specific relationship with its associated reference orientation, this is an inherent problem of the inverse light scattering pattern. Fortunately, for cases where the uncertainty in our signal measurements is low, we can easily predict the orientation by utilising computational techniques such as neural networks.

We utilised a python machine learning program called scikit-learn to build a neural network for identifying the dimer's orientation from its light scattering signal. The network was trained by generating a database of random orientation vectors, calculating the light scattering signals, and then having the network estimate the probability of the signal coming from our dimer in that corresponding reference orientation. After making its estimation the network's loss function was evaluated and used to improve the estimation, we trained the network until the improvement in the loss function was less than 0.0001. 
%
The estimation provided by the neural network can be improved further by accounting for any prior information we know about the dimer. We therefore, utilise Bayesian inference to update our neural networks estimation.

\begin{align}
  \label{eq:bayes}
  p(\hat{\bf n}_\alpha| y_k(\hat{\bf s}))
  &=
    \frac{p(y_k(\hat{\bf s})|\hat{\bf n}_\alpha)
    p(\hat{\bf n}_\alpha)}{p(y_k(\hat{\bf s}))}
\end{align}
where $p(\hat{\bf n}_\alpha)$ and $p(y_1, y_2, y_3)$ are the prior estimates of the distributions of particle orientations and instantaneous signals, respectively.
%
Without any prior evidence we assume that the orientation priori of our dimer
$p(\hat{\bf n}_{\alpha})$ is uniform. Knowledge of the dimer from previous
measurement can be used to inform our estimate of $p(\hat{\bf n}_{\alpha})$, covered in Section~\ref{sec:test}.  The latter prior $p(y)$ is the probability of measuring a signal ($y_1$, $y_2$, $y_3$).  This is given by taking the discrete integral over collection of reference orientations.
\begin{align}
  p(y_1, y_2, y_3)
  =
  \sum_{\alpha=1}^{n_{\rm ref}}
  p(y_1, y_2, y_3|\hat{\bf n}_\alpha)
  p(\hat{\bf n}_\alpha)
\end{align}
From \eqref{eq:bayes} we get a mass probability distribution that denotes the
probability that our dimer is in orientation $\hat{\bf n}_{\alpha}$ given a
signal ($y_1$, $y_2$, $y_3$). In order to evaluate the accuracy of our method, we tested it using a Brownian dynamics simulation of the dimer (see Section~\ref{sec:brownian}).

@@

As can be seen from Figure~\ref{fig:epsilon}, the inclusion of a signal noise quickly leads to a decrease in our model's performance. This is due to an inherent feature of the inverse scattering problem; that being that a two distinct regions of orientation space will become heavily intertwined in the intensity space resulting in poor predictions from our neural network. Fortunately, we can alleviate this by increasing the weighting factor $\beta$ that reflects the degree of freedom our dimer has within the trap. We repeated the model at for the above signal errors, but increasing $\beta$ until the model best fit our dimer's motion

@@

>>>>

\begin{highlights}
\item Asymmetric dimers undergo a full inversion in the presence of an potential well. This only occurs if the size ratio is greater than 1:2.  
\item Orientations are classified by a combination of neural networks and Bayesian inference. 
\item Signal error has a significant effect on the performance of the neural network, this can be countered via biasing of the priori distribution. 
\item Future applications in continuos processes where particle size is changing with time are being considered.  
\end{highlights}

@@

Since the invention of optical tweezers in the late 1980's, they have
found applications in experiments ranging from single molecule
biophysics \cite{Bustamante2021Biophysics} to those that test the
fundamental assumptions of quantum mechanics \cite{yin2013large}
thanks mainly to the ability of the tweezer to transduce and detect
forces on the order of a few piconewtons. \textit{Further
  characterisation of the trapped entity would provide useful
  information, as explored in for example} metrological measurements
\cite{arita2020coherent} and colloidal aggregation
\cite{burns1990optical}. Chemical characterisation of the trapped
entities has been achieved using spectroscopic techniques such as
Raman Scattering \cite{gupta2014raman}, while dynamical
characterisation has been demonstrated using a Quadrant Photo
Detector, by following the centre-of-mass Brownian motion of the
trapped entity \cite{friedrich2012tuning} and measuring rotation of
the centre-of-mass \cite{yifat2021facile}.

@@

\subsection{Brownian Dynamics Simulation}
\label{sec:brownian}

Brownian dynamics simulations of the motion of a silica dimer trapped
in a highly focused Gaussian beam were performed using software
package Brownian~OT, developed by Fung~\textit{et~al}.\
\cite{Vigilante2020Brownian_OT} to simulate the motion of a asymmetric
dimer within an optical trap.  Brownian~OT combines MSTM
\cite{Mishchenko1996MSTM} and ``Optical Tweezer Toolbox''
(\textit{ott}) \cite{Lenton2020} to determine the optical forces and
torques on arbitrary clusters of spheres using the T-Matrix method.
%MSTM
%computes the T-matrix of the dimer, and then passes the T-matrix to
%\textit{ott} to compute the optical forces.
%The benefit of this approach is that MSTM can accurately compute the
%T-matrix for any arbitrary collection of spheres once, and then
%\textit{ott} can calculate the optical forces and torques.
%
The diffusion tensor of the dimer taken from Ref.~\citenum{Nir_Acrivos_1973}.
%Ref.~\citenum{Farsund1996}.
%
%The total displacement is the sum of the optical and Brownian
%contributions.
%
Brownian OT returns the dimer's motion as a \textit{csv} file
containing for each time step the $[x,y,z]$ position of the dimer's
centre of mass relative to the trap focus, and the orientation
represented as a quaternion.

@@

Our goal is to determine the orientation of our trapper dimer based on
the scattering measurements $I(\hat{n}, \theta_k)$ collected from
three detection fibres situated at $\theta_1$, $\theta_2$, $\theta_3$,
and $\theta_4$.  The dimer's orientation is discretized into
$n_{ref}=30$ reference orientations $\hat{\bf n}_{\alpha}$ that are
evenly distributed on a unit sphere \cite{Rey2006} (see
Figure~\ref{fig:dimer}~b).  The spacing between two neighbouring
reference orientations is roughly $0.895$~radians.

Using MSTM, we compute the raw intensities that would be collected
from the dimer in each reference orientation
$I(\hat{\bf n}_{\alpha}, \theta_k)$. We noticed that keeping all four
angles in the same plane resulted in identical signals, to counter
this we projected $\theta_4$ out of the horizontal plane and into the
vertical plane ($\phi = 90$) resulting in unique signals for each
reference orientation.
%
These raw intensities are normalized according to
\begin{align}
\label{eq:scale}
  y_k(\hat{\bf n}_\alpha)
  = 
  \frac{I(\hat{\bf n}_\alpha, \theta_k) - \langle I(\hat{\bf n},\theta_k) \rangle } 
  {\langle I^2(\hat{\bf n},\theta_k) \rangle -\langle I(\hat{\bf n}, \theta_k)\rangle^2}
\end{align}
where the denominator is simply the standard deviation for our values
of $I(\hat{\bf n}, \theta_k)$.  The reference orientations, raw
intensities, and scaled signals are given in Tables~\ref{tab:A1} and
\ref{tab:A2}.  The distribution of collected signals bares no specific
relationship with its associated reference orientation, this is an
inherent problem of the inverse light scattering pattern.

%Fortunately, for cases where the uncertainty in our signal
%measurements is low, we can easily predict the orientation by
%utilising computational techniques such as neural networks.

The inverse mapping from the scaled measured signal to a particular
reference orientation of the dimer was fit using a multi-layer
perceptron neural network with four hidden layers, each with 100
neurons logistic activation function.  This was implemented in Python
using the scikit-learn \cite{Pedregosa_etal_2011} library.
%to build a neural network for
%identifying the dimer's orientation from its light scattering
%signal.
The neural network was trained using the stochastic optimizer Adam
\cite{Kingma_Ba_014} on a data set that was generated by randomly
choosing $10^4$ orientations and calculating the corresponding scaled
signals from light scattering.
%
%
%and then having the network estimate the probability of the
%signal coming from our dimer in that corresponding reference
%orientation.  After making its estimation the network's loss function
%was evaluated and used to improve the estimation, we trained the
%network until the improvement in the loss function was less than
%$10^{-4}$.

The estimate provided by the neural network can be improved further by
accounting for any prior information we know about the dimer. We
therefore, utilise Bayesian inference to update our neural networks
estimation.
\begin{align}
  \label{eq:bayes}
  p(\hat{\bf n}_\alpha| y_k(\hat{\bf s}))
  &=
    \frac{p(y_k(\hat{\bf s})|\hat{\bf n}_\alpha)
    p(\hat{\bf n}_\alpha)}{p(y_k(\hat{\bf s}))}
\end{align}
where $p(\hat{\bf n}_\alpha)$ and $p(y_1, y_2, y_3)$ are the prior estimates of the distributions of particle orientations and instantaneous signals, respectively.
%
Without any prior evidence we assume that the orientation priori of our dimer
$p(\hat{\bf n}_{\alpha})$ is uniform. Knowledge of the dimer from previous
measurement can be used to inform our estimate of $p(\hat{\bf n}_{\alpha})$, covered in Section~\ref{sec:test}.  The latter prior $p(y)$ is the probability of measuring a signal ($y_1$, $y_2$, $y_3$).  This is given by taking the discrete integral over collection of reference orientations.
\begin{align}
  p(y_1, y_2, y_3)
  =
  \sum_{\alpha=1}^{n_{\rm ref}}
  p(y_1, y_2, y_3|\hat{\bf n}_\alpha)
  p(\hat{\bf n}_\alpha)
\end{align}
From \eqref{eq:bayes} we get a mass probability distribution that denotes the
probability that our dimer is in orientation $\hat{\bf n}_{\alpha}$ given a
signal ($y_1$, $y_2$, $y_3$). In order to evaluate the accuracy of our method, we tested it using a Brownian dynamics simulation of the dimer (see Section~\ref{sec:brownian}).

@@

As can be seen from Figure~\ref{fig:epsilon}, the inclusion of a signal noise quickly leads to a decrease in our model's performance. This is due to an inherent feature of the inverse scattering problem; that being that a two distinct regions of orientation space will become heavily intertwined in the intensity space despite remaining continuous. As the signal error increases our neural network will map the signal to the incorrect reference orientation.
%
Fortunately, we can alleviate this by increasing the weighting factor $\beta$ that reflects the degree of freedom our dimer has within the trap. We repeated the model at for the above signal errors, but increasing $\beta$ until the model best fit our dimer's motion

@@

Additionally, we found that the due to the inverse light scattering problem that physical space regions can be continuously mapped to a space in the intensity region but with a loss of shape and size. In the future, we believe that we can improve our estimation by utilising a greater number of reference orientations and by optimising our choice of detection angles to better discretise our intensity regions.
